```{r setup, include = FALSE}
include = FALSE
echo = FALSE
opts_chunk$set(echo = TRUE
	       , warning = FALSE
	       , message = FALSE
	       , dpi = 300)
options(digits = 3)

tyxtable <- function(x, trim = 2) {
tmp <- textConnection(capture.output(xtable(x), file = NULL))
tmp <- readLines(tmp)
tmp <- tmp[-1:(-1 * trim)]
writeLines(tmp)
}
```

```{r packages, include = FALSE}
library(knitr)
library(dplyr)
```

STAT 293, Term 2 2014

Duncan Garmonsway 30020831

# Assignment 1

1. In the following groups of expressions ((1) and (2)) indicate which values will be the same as another.
   1) (a) and (d) are the same because both represent the sum of all rows of the $j$\textsuperscript{th} group.
   2) (a) and (d) are the same.  All other pairs imply that the sum of the squares is the same as the square of the sums, which is false, e.g. $y_1 = 1, y_2 = -1, \sum y_i^2 = 2, \left( \sum y_i \right)^2 = 0$.
2. Expectations and Variances
   a) If $E(X_i) = \mu_i$ for $i = 1, ..., n$ find $E(\bar{X})$
\begin{align*}
E(\bar{X}) &= E \left( \frac{1}{k} \sum_{i = 1}^{k} X_i \right) \\
&= \frac{1}{k} E \left( \sum_{i = 1}^{k} X_i \right) \\
&= \frac{1}{k} \sum_{i = 1}^{k} E(X_i) \\
&= \frac{1}{k} \sum_{i = 1}^{k} \mu_i \\
\end{align*}
\newpage
   b) If $Var(X_i) = \sigma^2$ for $i = 1, ..., n$ find $Var(\bar{X})$
\begin{align*}
Var(\bar{X}) &= Var\left(\frac{1}{k} \sum_{i = 1}^{k} X_i \right) \\
&= Var\left( \sum_{i = 1}^{k} \frac{1}{k} X_i \right) \\
&= \sum_{i = 1}^{k} \left( \frac{1}{k} \right)^2 Var(X_i) + \sum_{i = 1}^{k} \sum_{j = 1}^{k} Cov(X_i, X_j) \\
&= \sum_{i = 1}^{k} \left( \frac{1}{k} \right)^2 \sigma^2 + \sum_{i = 1}^{k} \sum_{j = 1}^{k} Cov(X_i, X_j) \\
&\text{ we stop here unless we can assume independence } \\
&= \sum_{i = 1}^{k} \left( \frac{1}{k} \right)^2 \sigma^2 = \frac{1}{k^2} \times k\sigma^2 = \frac{\sigma^2}{k}
\end{align*}
   c) Given the model: $Y_{ij} = \mu + \alpha_i + \epsilon_{ij}$ with $\mu$ and $\alpha_i$ fixed and $\epsilon_{ij} \overset{\mathrm{iid}}{\sim} \mathrm{N}(0, \sigma^2)$, what is $E(Y_{ij})$ and $Var(Y_{ij})$?
\begin{align*}
E(Y_{ij}) &= \mu + \alpha_i \\
Var(Y_{ij}) &= E\left[(Y_{ij} - E(Y_{ij}))^2\right] \\
&\text{ substitute } E(Y_{ij}) = \mu + \alpha_i \text{ and } Y_{ij} = \mu + \alpha_i + \epsilon_{ij} \\
&= E\left[(\mu + \alpha_i + \epsilon_{ij} - (\mu + \alpha_i))^2\right] \\
&= E\left(\epsilon_{ij}^{2}\right) \\
&= \sigma^2
\end{align*}

3.  Study.
   a) $Y_{ij} = \mu + \alpha_i + \epsilon_{ij}$
$Y_{ij}$ is a random variable for the improvement in total symptoms score for child $j$ given treatment $i$ (1 = control, 2\nobreakspace =\nobreakspace dextromethorphan, 3 = honey).
\newline\newline
$\sum_{i = 1}^{3} \alpha_i = 0$ (The effects sum to zero)
\newline\newline
$\epsilon_{ij} \overset{\mathrm{iid}}{\sim} \mathrm{N}(0, \sigma^2)$ (The errors are from identical standard normal distributions, independently)
\newpage
   b) Boxplot.

```{r question_03_b, fig.width = 5, fig.height = 3.5, units = "in", fig.cap = ""}
x <- read.table("honeycough.txt", header = TRUE)
levels(x$Treatment) <- c("control", "dextromethorphan", "honey")
boxplot(Score ~ Treatment, 
	data = x, main = "Improvement in total symptoms score",
	xlab = "Treatment", ylab = "Improvement")
```

   c) Table of mean and variance of improvement scores per treatment group.

```{r question_03_c, results = "asis"}
library(dplyr) # Using the dplyr package by Hadley Wickham
library(knitr) # and the knitr package by Yihui Xie
x_summary <- x %>% group_by(Treatment) %>% 
summarize(Mean = mean(Score), Variance = var(Score))
kable(x_summary)
```

   Note that the published figures were $8.39$ (dextromethorphan) and $6.41$ (control).

   d) Overall mean and estimate of $\alpha_i$.

```{r question_03_d}
# overall mean
mean(x$Score)

# estimate of effect of each treatment group in one line of base R
tapply(x$Score, x$Treatment, function(y) {mean(y) - mean(x$Score)})
```

   e) Model fitting with `lm` and a table of coefficient estimates.

```{r question_03_e, results = "asis"}
x_lm_summary <- summary(lm(Score ~ Treatment, data = x))
x_table <- x_lm_summary$coefficients[, 1, drop = FALSE]
x_table <- x_lm_summary$coefficients
rownames(x_table) <- c("Intercept ($\\mu$)",
		       "Effect of dextromethorphan ($\\alpha_2$)",
		       "Effect of honey ($\\alpha_3$)")
kable(x_table)
```

   Honey effected a larger average improvement in cough symptoms, compared with the control, than dextromethorphan (both significant at the 95% level).  Since the intercept is positive, the children may have been recovering regardless of the treatment.
