```{r setup, include = FALSE}
include = FALSE
echo = FALSE
opts_chunk$set(echo = TRUE
	       , warning = FALSE
	       , message = FALSE
	       , dpi = 300)
options(digits = 3)

tyxtable <- function(x, trim = 2) {
tmp <- textConnection(capture.output(xtable(x), file = NULL))
tmp <- readLines(tmp)
tmp <- tmp[-1:(-1 * trim)]
writeLines(tmp)
}
```

```{r packages, include = FALSE}
library(knitr)
library(dplyr)
```

STAT 293, Term 2 2014

Duncan Garmonsway 30020831

# Assignment 1

1. In the following groups of expressions ((1) and (2)) indicate which values will be the same as another.
   1) (a) and (d) are the same because both represent the sum of all rows of the j\textsuperscript{th} group.
   2) (a) and (d) are the same.  All other pairs imply that the sum of the squares is the same as the square of the sums, which is false, e.g. $y_1 = 1, y_2 = -1, \sum y_i^2 = 2, \left( \sum y_i \right)^2 = 0$.
2. Expectations and Variances
   a) If $E(X_i) = \mu_i$ for $i = 1, ..., n$ find $E(\bar{X})$
\begin{align*}
E(\bar{X}) &= E \left( \frac{X_1 + X_2 + ... + X_n}{n} \right) \\
&= E(X_1/n) + E(X_2/n) + ... + E(X_n/n) \\
&= \frac{1}{n}E(X_1) + ... + \frac{1}{n}E(X_n) \\
&= \frac{1}{n} \sum_{i = 1}^{n} E(X_i) \\
&= \frac{1}{n}n\mu \\
&= \mu
\end{align*}
   b) If $Var(X_i) = \sigma^2$ for $i = 1, ..., n$ find $Var(\bar{X})$
\begin{align*}
Var(\bar{X}) &= Var\left(\frac{1}{n} \sum_{i = 1}^{n} X_i \right) \\
&= Var\left( \sum_{i = 1}^{n} \frac{1}{n} X_i \right) \\
&= \sum_{i = 1}^{n} \left( \frac{1}{n} \right)^2 Var(X_i) + \sum_{i = 1}^{n} \sum_{j = 1}^{n} Cov(X_i, X_j) \\
&= \sum_{i = 1}^{n} \left( \frac{1}{n} \right)^2 \sigma^2 \text{ since we ignore covariance in this paper} \\
&= \frac{1}{n^2} \times n\sigma^2 \\
&= \frac{\sigma^2}{n}
\end{align*}
   c) Given the model: $Y_{ij} = \mu + \alpha_i + \epsilon_{ij}$ with $\mu$ and $\alpha_i$ fixed and $\epsilon_{ij} \overset{iid}{\sim} \mathrm{N}(0, \sigma^2)$, what is $E(Y_ij)$ and $Var(Y_ij)$?
   \newline \newline
\begin{align*}
E(Y_{ij}) &= \mu + \alpha_i \\
Var(Y_{ij}) &= E\left((Y_{ij} - E(Y_ij))^2\right) \\
&\text{ substitute } E(Y_{ij}) = \mu + \alpha_i \text{ and } Y_{ij} = \mu + \alpha_i + \epsilon_{ij} \\
&= E\left((\mu + \alpha_i + \epsilon_{ij} - (\mu + \alpha_i))^2\right) \\
&= E\left(\epsilon_{ij}^2\right) \\
&= \sigma^2
\end{align*}

3.  Study.
   a) $Y_{ij} = \mu + \alpha_i + \epsilon_{ij}$
$Y_{ij}$ is a random variable for the improvement in total symptoms score for child $j$ given treatment $i$ (1 = control, 2 = dextromethorpan, 3 = honey).
\newline\newline
$\sum_{i = 1}^{3} \alpha_i = 0$
\newline\newline
$\epsilon_{ij} \overset{iid}{\sim} \mathrm{N}(0, \sigma^2)$
   b) Boxplot.

```{r question_03_b, fig.width = 5, fig.height = 3, units = "in", fig.cap = ""}
x <- read.table("honeycough.txt", header = TRUE)
levels(x$treatment) <- c("control", "dextromethorpan", "honey")
boxplot(score ~ treatment, 
	data = x,
	main = "Improvement in total symptoms score",
	xlab = "Treatment",
	ylab = "Improvement")
```

   c) Table of mean and variance of improvement scores per treatment group.

```{r question_03_c, results = "asis"}
# Using the dplyr package by Hadley Wickham
# and the knitr package by Yihui Xie
library(dplyr)
library(knitr)
x_summary <- x %>% 
group_by(treatment) %>% 
summarize(mean = mean(score), variance = var(score))

kable(x_summary)
```

   Note that the published figures were $8.39$ (dextromethorpan) and $6.41$ (control).

   d) Overall mean and estimate of $\alpha_i$.

```{r question_03_d}
# overall mean
mean(x$score)

# estimate of effect of each treatment group in one line of base R
tapply(x$score, x$treatment, function(y) {mean(y) - mean(x$score)})
```

   e) Model fitting with `lm` and a table of coefficient estimates.

```{r question_03_e, results = "asis"}
x_lm_summary <- summary(lm(score ~ treatment, data = x))
x_table <- x_lm_summary$coefficients[, 1, drop = FALSE]
x_table <- x_lm_summary$coefficients
rownames(x_table) <- c("intercept (mu)",
		       "effect of dextromethorpan (alpha 2)",
		       "effect of honey (alpha 3)")
kable(x_table)
```

   Honey effected the largest average improvement in cough symptoms, followed by dextromethorpan (both significant at 95% level).  Since the intercept is positive, the children may have been recovering regardless of the treatment.
