```{r setup, include = FALSE}
include = FALSE
echo = FALSE
opts_chunk$set(echo = TRUE
	       , warning = FALSE
	       , message = FALSE
	       , dpi = 300)
options(digits = 3)

tyxtable <- function(x, trim = 2) {
tmp <- textConnection(capture.output(xtable(x), file = NULL))
tmp <- readLines(tmp)
tmp <- tmp[-1:(-1 * trim)]
writeLines(tmp)
}
```

```{r packages, include = FALSE}
library(xtable) # to print matrix in latex
```

STAT 293, Term 2 2014

Duncan Garmonsway 30020831

# Tutorial 1

1. Complete
   a) Exercise 1
   b) Exercise 2
   c) Exercise 3 \newline\newline
   in the **Introduction to R** handout (given out during first lecture and available on **R** page of course website).

   \setcounter{enumi}{1}
   a) $\text{Fahrenheit} = 9/5 \times \text{Celsius} + 32$.  The boiling point of water is $100$ degrees Celsius.  Write a fragment of **R** code to determine the boiling point of water in degrees Fahrenheit.

```{r excersise_1}
celsius_to_fahrenheit <- function(degrees) {
    9/5 * degrees + 32
}

celsius_to_fahrenheit(100)
```

   b) Write a fragment of **R** code to enter a vector of temperature readings in degrees Fahrenheit, convert these to Celsius scale using an appropriate formula and print out the result.

```{r exercise_2}
celsius_to_fahrenheit(0:9)
```

   c) Find all solutions to the quadratic equation $2x^2 - 10x + 8 = 0$

```{r exercise_3}
coeff_a <- 2
coeff_b <- -10
coeff_c <- 8

(0 - coeff_b + sqrt(coeff_b^2 - 4 * coeff_a * coeff_c)) / (2 * coeff_a)
(0 - coeff_b - sqrt(coeff_b^2 - 4 * coeff_a * coeff_c)) / (2 * coeff_a)
```

2. Five randomly selected STAT293 students were each given a pedometer and asked to record the number of steps walked during the six-hour period between 12 noon and 6 pm on three successive days.  The data they submitted are presented in array format below where row $i$ corresponds to individual $i$ for $i = 1, 2, ..., 5$; column $j$ represents day $j$ for $j = 1, 2, 3$; and each observation is presented by $y_ij$.
\begin{align*}
```{r matrix, echo = FALSE, results = "asis"}
x <- matrix(c(1053, 3029, 2001
	      , 5745, 5095, 6790
	      , 2569, 3975, 2011
	      , 1500, 4500, 5000
	      , 6397, 2056, 7309), 5, 3, byrow = TRUE)
xprint <- xtable(x, align = rep("", ncol(x) + 1))
print(xprint
      , floating = FALSE
      , tabular.environment = "bmatrix"
      , hline.after = NULL
      , include.rownames = FALSE
      , include.colnames = FALSE
      , comment = FALSE)
```
\end{align*}
Use **R** to compute the value of each of the following terms and state what each value represents (`x` is the variable containing the matrix).
a)  $y_{1.}$ The sum of row 1 is ```sum(x[1, ])```$= `r sum(x[1, ])`$.
b)  $\overline{y}_{3.}$ The mean of row 3 is ```mean(x[3, ])```$= `r mean(x[3, ])`$.
c)  $\overline{y}_{.3}$ The mean of column 3 is ```mean(x[, 3])```$= `r mean(x[, 3])`$.
d)  $\overline{y}_{..}$ The mean of the matrix is ```mean(x)```$= `r mean(x)`$.

3. Suppose there are 5 random variables: $X_1, X_2, ..., X_5$, each with mean $\mu$ and variance $\sigma^2$ and that each of these variables is independent of the other.  Find:
   a) $E\left( \sum_{i = 1}^{5} X_i \right)$
\begin{align*}
& = E(X_1 + X_2 + ... + X_5) \\
& = E(X_1) + E(X_2) + ... + E (X_5) \\
& \text{Since the mean of each variable is } \mu \text{ then} \\
& = 5\mu 
\end{align*}
   b) $E(\overline{X}) = \frac{1}{5} (\overline{X}_1 + \overline{X}_2 + ... + \overline{X}_5) = \frac{1}{5}(5\mu) = \mu$
   \newline
   c) $\text{Var} \left( \sum_{i = 1}^{n} X_i \right) = \sum_{i = 1}^{n} \text{Var}(X_i) \newline\text{ (since the covariance of independent variables is }0\text{) } = 5\sigma$
   \newline
   d) $\text{Var}(\overline{X})$
