```{r setup, include = FALSE}
include = FALSE
echo = FALSE
opts_chunk$set(echo = TRUE
	       , warning = FALSE
	       , message = FALSE
	       , dpi = 300)
options(digits = 3)
```

```{r packages, include = FALSE}
```

STAT 293, Term 2 2014

Duncan Garmonsway 30020831

# Assignment 2

1. a)
\begin{align*}
E \left( \bar{Y_{i.}} - \bar{Y_{i;.}} \right) \\
&=E \left( \frac{1}{n_i} \sum_{j=1}^{n_i} Y_{ij} - \frac{1}{n_{i'}} \sum_{j=1}^{n_{i'}} Y_{ij} \right) \\
&=E \left( \frac{1}{n_i} \sum_{j=1}^{n_i} \left( \mu + \alpha_i + \epsilon_{ij} \right) - \frac{1}{n_{i'}} \sum_{j=1}^{n_i'} \left( \mu + \alpha_{i'} + \epsilon_{i'j} \right) \right) \\
&=E \left( \mu + \alpha_i + \epsilon_{ij} - \mu - \alpha_{i'} - \epsilon_{i'j} \right) \\
&=E \left( \alpha_i - \alpha_{i'} \right) + E(\epsilon_{ij}) - E(\epsilon_{i'j}) \\
&=E \left( \alpha_i - \alpha_{i'} \right) + 0 - 0 \\
&=E \left( \alpha_i - \alpha_{i'} \right)
\end{align*}
   b)
\begin{align*}
Var(\bar{Y_{i.}} - \bar{Y_{i'.}}) \\
&=Var(\bar{Y_{i.}}) + Var(\bar{-Y_{i'.}})) \\
&=(1)^2 Var(\bar{Y_{i.}}) + (-1)^2 Var(\bar{Y_{i'.}})) \\
&=Var(\bar{Y_{i.}}) + Var(\bar{Y_{i'.}}) \\
&=\frac{\sigma^2}{n_i} + \frac{\sigma^2}{n_{i'}} \text{ (p.12 of the notes)} \\
&=\sigma^2(\frac{1}{n_i} + \frac{1}{n_{i'}})
\end{align*}

2. a)
\begin{align*}
5 - 1 &= 4 \text{ so } X_1 \sim \chi^2_{4} \\
12 - 7 &= 5 \text{ so } X_2 \sim \chi^2_{5} \\
15 - 3 &= 12 \text{ so } X_3 \sim \chi^2_{12}
\end{align*}

   b) $W \sim \chi^2_{(5 + 6 + 13 = 24)}$ 

   c) $W \sim \mathrm{F}_{(6, 13)}$

3. a)
```{r question03a, include = TRUE}
Ybar_i <- c(4.054, 3.9977, 4.8130)
s_i <- c(.7791, .8413, .6615)
n_i <- c(10, 10, 30)

n <- sum(n_i)
k <- 3

(SSE <- sum((n_i - 1) * s_i^2))

(SST <- sum(n_i * Ybar_i^2) - (sum(n_i * Ybar_i)^2 / sum(n_i)))
```

   b) $SSE / \sigma^2 \sim \chi^2_{(40 - 3 = 37)} \text{ and } SST / \sigma^2 \sim \chi^2_{(3 - 1 = 2)}$

   c)
```{r question03c, include = TRUE}
(MSE <- SSE / (n - k))

(MST <- SST / (k - 1))

(Fstat <- MST / MSE)

(pval <- 1 - pf(Fstat, k - 1, n - k))
```
Since $p < 0.02$ there is sufficient evidence of differences among mean residue amounts at the 2% significance level, assuming all ANOVA assumptions have been met.
