```{r setup, include = FALSE}
include = FALSE
echo = FALSE
opts_chunk$set(echo = TRUE
	       , warning = FALSE
	       , message = FALSE
	       , dpi = 300)
options(digits = 3)
```

```{r packages, include = FALSE}
library(knitr)
```

STAT 293, Term 2 2014

Duncan Garmonsway 30020831

# Assignment 3

1.  a) $\mathcal{H}_0:$ There was no difference of mean score in the word completion task between treatment groups.  $\mathcal{H}_A:$ The mean scores differed between treatment groups.

   b)
```{r question1b}
alcohol <- read.csv("alcohol.csv")
(Fobs <- anova(lm(score ~ group, data = alcohol))$`F value`[1])
```
   c)
```{r question1c}
set.seed(3243) # reproducibility
permutations = 1000

Fnull <- rep(NA, permutations)
for (t in 1:1000) {
    # Randomly reorder the groups.
    group_reorder <- sample(alcohol$group)
    # Perform an anova of score against the reordered groups and
    # store the F statistic.
    Fnull[t] <- anova(lm(alcohol$score ~ group_reorder))$`F value`[1]
}
# proportion of Fnull values >= Fobs
(p <- sum(Fnull >= Fobs) / permutations)
```
   Since the $p$-value is $0$ from $1000$ permutations, we estimate that the true $p$-value is less than $0.001$ so we reject $\mathcal{H}_0$ at the 5% significance level; there was a significant difference of mean scores in the word completion task between different treatment groups.

   d)
```{r question3d}
(p_anova <- anova(lm(score ~ group, data = alcohol))$`Pr(>F)`[1])
```
   Since the $p$-value of the ANOVA is less than $0.001$ we reject $\mathcal{H}_0$ at the 5% significance level; there was a significant difference of mean scores in the word completion task between different treatment groups.

   e) If the ANOVA assumptions were met I would present the ANOVA conclusions as the ANOVA process is better known, but if the assumptions are not met then I would present the conclusions of the permutation test as they are more likely to be correct than ANOVA in such a case.

2. a)
```{r question2a, fig.height = 4}
factpoison <- read.table("factpoison.txt"
			 , header = TRUE
			 , colClasses = c("factor", "factor", "numeric"))
factpoison_lm <- lm(Time ~ Poison * Treatment, data = factpoison)
par(mfrow = c(1, 2))
plot(factpoison_lm, which = 1:2, ask = FALSE)

# Levene's test using the car package
car::leveneTest(factpoison_lm)$`Pr(>F)`[1]
```
   The assumption of constant variance is not met because there is a clear funneling pattern in the plot of residuals $vs$ fitted values, and because Levene's test is significant at the 5% level.  The assumption of normality of is not met because the q-q plot is far from straight.  Given these departures from the assumptions, I expect the conclusions from a normal theory ANOVA test to differ from those reached from a permutation test because the distribution of the F statistic under the null hypothesis may not be from the F distribution, so the $p$-value may be different.

   b)i)
```{r question2bi, results = "asis"}
kable(anova(factpoison_lm))
```

   j)
```{r question2bii}
# To install an archived package, find the direct link on CRAN and use devtools, e.g.
# devtools::install_url("http://cran.r-project.org/.../lmPerm_1.1-2.tar.gz")
library(lmPerm)
anova(lmp(Time ~ Poison * Treatment, data = factpoison))
# doesn't return an object = doesn't deserve to be on CRAN!
```

3. a) False; according to the documentation it will still use a random sample of permutations if the number of observations exceeds the "maxExact" parameter, which defaults to 10.

   b) False; `anova.lm()` defaults to Type I SS (sequential), whereas `anova.lmp()` defaults to type III.

   c) True

   d) True

   e) True
